Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments

### 2021-05-17 doc/paper.Rmd
em dashes

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwood2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwood2015].

@@

FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
There was no evidence for a cost of reproduction --- in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
However, the coefficients were always positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
This suggests that....<!--# not surew what to make of this lastsentence. Interpret for reader?   -->

>>>>

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwood2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
Plant populations may be particularly sensitive---an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwood2015].

@@

FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
There was no evidence for a cost of reproduction---in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
However, the coefficients were always positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
This suggests that....<!--# not surew what to make of this lastsentence. Interpret for reader?   -->
